\papertitle\footnote{Manuscript under review; see \url{http://geoffboeing.com/publications/multiscale-analysis-street-networks/}}

### 2018-02-14 draft/v3/main.tex
update figures and layout

\documentclass[Afour,sageh,times]{sagej}

\usepackage{graphicx}
\usepackage{geometry}
\usepackage{booktabs}
\usepackage{pdflscape}
\usepackage{setspace}
\usepackage[USenglish]{babel}
\usepackage{csquotes}
\usepackage{natbib}
\usepackage{url}

\usepackage{moreverb}
\usepackage[colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}
\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
		T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\def\volumeyear{2017}

\renewcommand{\cite}{\citep}
\doublespacing

\begin{document}
\title{A Multi-Scale Analysis of 27,000 Urban Street Networks}
\author{Author Redacted}
\date{March 2017}
\maketitle

\begin{abstract}

OpenStreetMap offers an underexplored source of worldwide geospatial data useful to urban researchers. This study uses the OSMnx software to automatically download and analyze 27,000 US street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---namely, every US city and town ($N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$). It presents wide-ranging empirical findings on US urban form and street network characteristics, emphasizing measures relevant to graph theory, urban design, and morphology such as structure, connectedness, density, centrality, and resilience. In the past, street network data acquisition and processing has been challenging and ad hoc. This study illustrates the use of OSMnx and OpenStreetMap to consistently conduct street network analysis with extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. These street networks and measures data sets have been shared in a public repository for other researchers to use.

\end{abstract}

@@

On May 20, 1862, Abraham Lincoln signed the Homestead Act into law, making land across the United States' Midwest and Great Plains available for free to applicants \cite{porterfield_homestead_2005}. Under its auspices over the next 70 years, the federal government distributed 10\% of the entire US landmass to private owners in the form of 1.6 million homesteads \cite{lee_kansas_1979, sherraden_inclusion_2005}. New towns with gridiron street networks sprang up rapidly across the Great Plains and Midwest, due to both the prevailing urban design paradigm of the day and the standardized rectilinear town plats used repeatedly to lay out instant new cities \cite{southworth_streets_1997}. Through path dependence, the spatial signatures of these land use laws, design paradigms, and planning instruments can still be seen today in these cities' urban forms and street networks. Cross-sectional analysis of American urban form can reveal these artifacts and histories through street networks at metropolitan, municipal, and neighborhood scales.

Network analysis is a natural approach to the study of cities as complex systems \cite{masucci_random_2009}. The empirical literature on street networks is growing ever richer, but suffers from some limitations---discussed in detail in \cite{redacted_osmnx:_2017} and summarized here. First, sample sizes tend to be fairly small due to data availability, gathering, and processing constraints: most studies in this literature that conduct topological or metric analyses tend to have sample sizes ranging around 10 to 50 networks \cite{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, strano_urban_2013}, which may limit the generalizability and interpretability of findings. Second, reproducibility has been difficult when the dozens of decisions that go into analysis---such as spatial extents, topological simplification and correction, definitions of nodes and edges, etc.---are ad hoc or only partly reported \citep[e.g.,][]{porta_network_2006, strano_urban_2013}. Third, and related to the first two, studies frequently oversimplify to planar or undirected primal graphs for tractability \cite[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009}, or use dual graphs despite the loss of geographic, metric information \cite{batty_network_2005, jiang_integration_2002, ratti_space_2004, crucitti_centrality_2006, crucitti_centrality_2006-1}.

This study addresses these limitations by conducting a morphological analysis of urban street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. In particular, it examines 27,000 urban street networks---represented as primal, nonplanar, weighted multidigraphs with possible self-loops---at multiple overlapping scales across the US, focusing on structure, connectedness, centrality, and resilience. It examines the street networks of every incorporated city and town (population size $N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$) in the US. To do so, it uses OSMnx---a new street network research toolkit \cite{redacted_osmnx:_2017}---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales. These street networks and measures data sets have been compiled and shared in a public repository for other researchers to repurpose at [redacted].

@@

A network (also called a \emph{graph} in mathematics) comprises a set of nodes connected to one another by a set of edges. Street networks can be conceptualized as primal, directed, nonplanar graphs. A \emph{primal} street network represents intersections as nodes and street segments as edges. A \emph{directed} network has directed edges: that is, edge $uv$ points one-way from node $u$ to node $v$, but there need not exist a reciprocal edge $vu$. A \emph{planar} network can be represented in two dimensions with its edges intersecting only at nodes \cite{viana_simplicity_2013, fischer_spatial_2014}. Most street networks are nonplanar---due to grade-separated expressways, overpasses, bridges, tunnels, etc.---but most quantitative studies of urban street networks represent them as planar \cite[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009, strano_urban_2013} for tractability because bridges and tunnels are uncommon in some cities. Planar graphs may reasonably model the street networks of old European town centers, but poorly model the street networks of modern autocentric cities like Los Angeles or Tokyo with many grade-separated expressways, bridges, and underpasses.

@@

This study uses OSMnx to download, construct, correct, analyze, and visualize street networks at metropolitan, municipal, and neighborhood scales. OSMnx is a Python-based research tool that easily downloads OpenStreetMap data for any place name, address, or polygon in the world, then constructs it into a spatially-embedded graph-theoretic object for analysis and visualization \cite{redacted_osmnx:_2017}. OpenStreetMap is a collaborative worldwide mapping project that makes its spatial data available via various APIs \cite{corcoran_analysing_2013, jokar_arsanjani_openstreetmap_2015}. These data are of high quality and compare favorably to CIA World Factbook estimates and US Census TIGER/Line data \cite{haklay_how_2010, over_generating_2010, zielstra_comparative_2011, maron_how_2015, wu_improving_2005, frizzelle_importance_2009}. In 2007, OpenStreetMap imported the TIGER/Line roads (2005 vintage) and since then, many community-led corrections and improvements have been made \cite{willis_openstreetmap_2008}. Many of these additions go beyond TIGER/Line's scope, including passageways between buildings, footpaths through parks, bike routes, and detailed feature attributes such as finer-grained street classifiers, speed limits, etc.

To define the study sites and their spatial boundaries, we use three sets of geometries. The first defines the metropolitan-scale study sites using the 2016 TIGER/Line shapefile of US census bureau urban areas. Each census-defined urban area comprises a set of tracts that meet a minimum density threshold \cite{u.s._census_bureau_2010_2010}. We retain only the \emph{urbanized areas} subset of these data (i.e., areas with greater than 50,000 population), discarding the small \emph{urban clusters} subset. The second set of geometries defines our municipal-scale study sites using 51 separate TIGER/Line shapefiles (again, 2016) of US census bureau \emph{places} within all 50 states plus DC. We discard the subset of \emph{census-designated places} (i.e., small unincorporated communities) in these data, while retaining every US city and town. The third set of geometries defines the neighborhood-scale study sites using 42 separate shapefiles from Zillow, a real estate database company. These shapefiles contain neighborhood boundaries for major cities in 41 states plus DC. This fairly new data set comprises nearly 7,000 neighborhoods, but as \citet{schernthanner_spatial_2016} point out, Zillow does not publish the methodology used to construct these boundaries. However, despite its newness it already has a track record in the academic literature: \citet{besbris_effect_2015} use Zillow boundaries to examine neighborhood stigma and \citet{albrecht_indicator_2014} use them to study neighborhood-level poverty in New York.

For each of these geometries, we use OSMnx to download the (drivable, public) street network within it, a process described in detail in \cite{redacted_osmnx:_2017} and summarized here. First OSMnx buffers each geometry by 0.5 km, then downloads the OpenStreetMap \enquote{nodes} and \enquote{ways} within this buffer. Next it constructs a street network from these data, corrects the topology, calculates street counts per node, then truncates the network to the original, desired polygon. OSMnx saves each of these networks to disk as GraphML and shapefiles. Finally, it calculates metric and topological measures for each network, summarized below. Such measures extend the toolkit commonly used in urban form studies \cite{talen_measuring_2003, ewing_travel_2010}.

@@

Brief descriptions of these OSMnx-calculated measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \cite{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. The \emph{average street segment length} is a linear proxy for block size and specifies the network's grain. \emph{Node density} divides the node count by the network's area, while \emph{intersection density} excludes dead-ends to represent the density of street junctions. \emph{Edge density} divides the total directed network length by area, while \emph{street density} does the same for an undirected representation of the graph (to not double-count bidirectional streets). \emph{Average circuity} measures the ratio of edge lengths to the great-circle distances between the nodes these edges connect, indicating the street pattern's curvilinearity (cf. [redacted]).

The network's \emph{average node degree} quantifies connectedness in terms of the average number of edges incident to its nodes. The \emph{average streets per node} adapts this for physical form rather than directed circulation. It measures the average number of physical streets that emanate from each node (i.e., intersection or dead-end). The distribution and proportion of streets per node characterize the type, pervasiveness, and spatial dispersal of network connectedness and dead-ends. Connectivity represents the fewest number of nodes or edges that will disconnect the network if they are removed and is thus an indicator of resilience. A network's \emph{average node connectivity} (ANC)---the mean number of internally node-disjoint paths between each pair of nodes---more usefully represents how many nodes must be removed on average to disconnect a randomly selected pair of nodes \cite{beineke_average_2002,dankelmann_bounds_2003}. Brittle points of vulnerability characterize networks with low average connectivity.

A node's \emph{clustering coefficient} represents the ratio between its neighbors' links and the maximum number of links that could exist between them \cite{jiang_topological_2004, opsahl_clustering_2009}. The weighted clustering coefficient weights this by edge length and the average clustering coefficient is the mean of the clustering coefficients of all the nodes. \emph{Betweenness centrality} evaluates how many of the network's shortest paths pass through some node (or edge) to indicate its importance \cite{barthelemy_betweenness_2004,huang_trajgraph:_2016, zhong_revealing_2017}. A network's \emph{maximum betweenness centrality} (MBC) measures the share of shortest paths that pass through the network's most important node: higher maximum betweenness centralities suggest networks more prone to inefficiency if this important choke point should fail. Finally, \emph{PageRank} ranks nodes based on the structure of incoming links and the rank of the source node \cite{brin_anatomy_1998, jiang_predicting_2008, agryzkov_algorithm_2012, chin_geographically_2015, gleich_pagerank_2015}.

@@

@@

@@

@@

\begin{figure*}
	\includegraphics[width=1\textwidth]{fig01.png}
	\caption{Intersection density per urbanized area (left) and average streets per node per urbanized area (right) from lowest (pale yellow) to highest (dark red), in the contiguous US}
	\label{fig:fig01}
\end{figure*}

\begin{figure*}
	\includegraphics[width=0.7\textwidth]{fig02.png}
	\caption{Distribution of node types in 9 urbanized areas, with number of streets emanating from the node on the $x$-axis and proportion of nodes of this type on the $y$-axis }
	\label{fig:fig02}
\end{figure*}

@@

\newgeometry{margin=1in}
\begin{landscape}

@@

\end{landscape}
\restoregeometry

@@

Table \ref{tab:measures_cities} presents summary statistics of street network characteristics across the entire data set of 19,655 cities and towns---every incorporated city and town in the US. Following recent work by \citet{barthelemy_modeling_2008} and \citet{strano_urban_2013}, we examine the relationship between the total street length $L$ and the number of nodes $n$ across different cities. The former proposed a model of city network evolution in which $L$ and $n$ scale nonlinearly as $n^{1/2}$, and the latter suggested that this relationship applies cross-sectionally, using a small empirical sample of ten European cities. However, the latter's small sample size may limit the generalizability of this finding. To investigate this empirically, we examine the relationship between $L$ and $n$ for 19,655 US cities and towns and instead find a strong \emph{linear} relationship ($r^{2}=0.98$), as depicted in Figure \ref{fig:fig03}. We also find a similar linear relationship at the metropolitan and neighborhood scales.

@@

Previous findings \citep[e.g.,][]{masucci_random_2009} also suggest street segment lengths in an urban network follow a power-law distribution. However, we find that these networks instead generally follow lognormal-style right-skewed distributions. This makes theoretical sense as most street networks are not truly scale-free: for example, a typical street network might comprise very few very long street segments (e.g., 1 km), more medium-length segments (e.g., 250 m), many short segments (e.g., 80 m), but very few very short segments (e.g., 10 m). To test this in a statistically rigorous way, we fit a set of candidate distributions to the street segment lengths of each US city/town. The candidate distributions comprise the lognormal, Gumbel, gamma, exponentiated Weibull, Fréchet, power-law, uniform, and exponential distributions. We then assess each fit with a Kolmogorov-Smirnov test, which tells us how closely each fitted distribution matches the observed data. Power-law distributions provide the best fit for fewer than 0.7\% of these cities. In contrast, the exponentiated Weibull distribution provides the best fit 56\% of the time, followed by the Gumbel (19\%), gamma (11\%), and lognormal (7\%) distributions.

An exception to this general pattern, of course, lies in consistently-sized orthogonal grids filling a city's incorporated spatial extents. Such distributions are extremely peaked around a single value: the linear length of a grid block. We find that such cities are not uncommon in the US, particularly between the Mississippi River and the Rocky Mountains: the Great Plains states are characterized by a unique street network form that is both orthogonal and reasonably dense. The \emph{former} is partly the result of topography (flat terrain that allows idealized grids) and design history (rapid platting and development during the late nineteenth century) that favor orthogonal grids, as discussed earlier. The \emph{latter} results from the fact that most towns across the Great Plains exhibit minimal suburban sprawl. Thus, municipal boundaries snugly embrace the gridlike street network, without extending to accommodate a vast peripheral belt of twentieth century sprawl, circuity, and \enquote{loops and lollipops} \cite{southworth_streets_1997} that characterizes cities in e.g. California that were settled in the same era but later subjected to substantial suburbanization.

\begin{figure*}
	\includegraphics[width=\textwidth]{fig03.png}
	\caption{The linear relationship between total street length and number of nodes in the street networks of 6,857 US neighborhoods (left) and 19,655 US cities and towns (right).}
	\label{fig:fig03}
\end{figure*}

\begin{figure*}
	\includegraphics[width=\textwidth]{fig04.png}
	\caption{Left: contiguous US states by median of mean streets per node in municipal street networks, from lowest (pale yellow) to highest (dark red). Right: contiguous US states by median of mean street segment length in municipal street networks, from longest (pale yellow) to shortest (dark red).}
	\label{fig:fig04}
\end{figure*}

\newgeometry{margin=2cm}

@@

@@

\restoregeometry

@@

@@

\begin{figure*}
	\includegraphics[width=1\textwidth]{fig05.png}
	\caption{Square-mile comparisons of central cities and their suburbs. Left: top, downtown Philadelphia, PA; bottom, its suburb, King of Prussia. Middle: top, downtown Portland, OR; bottom, its suburb, Beaverton. Right: top, downtown San Francisco, CA; bottom, its suburb, Concord.}
	\label{fig:fig05}
\end{figure*}

To illustrate these morphologies, Figure \ref{fig:fig05} compares one square mile of the centers of Philadelphia, Portland, and San Francisco with one square mile of each of their suburbs. The connectedness and fine-grain of the central cities are clear, as are the disconnectivity and coarse-grain of their suburbs. In fact, the suburbs have more in common with one another---despite being hundreds or thousands of miles apart---than they do with their central city neighbors, suggesting that land use and an era's prevailing design paradigm is paramount to geographical localism and regional context. The top row of Figure \ref{fig:fig05} represents an era of planning and development that preceded the automobile, while the bottom row reflects the exclusionary zoning and mid--late twentieth century era of automobility in residential suburb design---namely the \enquote{loops and lollipops} and the \enquote{lollipops on a stick} design patterns \cite{southworth_streets_1997}.

>>>>

%-------------------------------------------------------------------------------------------
% enter the author and paper information below:
%-------------------------------------------------------------------------------------------
\newcommand{\myname}{Geoff Boeing}
\newcommand{\myemail}{gboeing@berkeley.edu}
\newcommand{\myaffiliation}{Department of City and Regional Planning\\University of California, Berkeley}
\newcommand{\paperdate}{February 2018}
\newcommand{\papertitle}{A Multi-Scale Analysis of 27,000 Urban Street Networks}
\newcommand{\papercitation}{Boeing, G. 2018. \papertitle. Under review at \emph{Journal Name}.}
\newcommand{\paperkeywords}{Urban Planning, Transportation, Data Science, Street Networks, Python, OpenStreetMap, Graph Theory, History}

%-------------------------------------------------------------------------------------------
% begin preamble setup
%-------------------------------------------------------------------------------------------
\RequirePackage[l2tabu,orthodox]{nag}
\documentclass[11pt]{article}

% import these packages first, in order
\usepackage[utf8]{inputenc}
\usepackage[strict,autostyle]{csquotes}
\usepackage[USenglish]{babel}

% import everything else
\usepackage{authblk}
\usepackage{booktabs}
\usepackage{crimson}
\usepackage{datetime}
\usepackage[final]{draftwatermark}
\usepackage{endnotes}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{microtype}
\usepackage{multicol}
\usepackage{natbib}
\usepackage{rotating}
\usepackage{setspace}
\usepackage{url}

% location of figure files
\graphicspath{{./figures/}}

% configure the page layout
\geometry{
	paper=letterpaper,         % paper size
	top=3cm,                   % margin sizes
	bottom=3cm,
	left=2.5cm,
	right=2.5cm}
\setlength{\columnsep}{0.75cm} % space between text columns
\setstretch{1.00}              % line spacing
\clubpenalty=10000             % prevent orphans
\widowpenalty=10000            % prevent widows
\let\footnote=\endnote         % turn footnotes into endnotes

% configure pdf metadata and link handling
\hypersetup{
	pdfauthor={\myname},
	pdftitle={\papertitle},
	pdfsubject={\papertitle},
	pdfkeywords={\paperkeywords},
	pdffitwindow=true,         % window fit to page when opened
	breaklinks=true,           % break links that overflow horizontally
	hidelinks=true             % remove link color and border
}

% configure watermark appearance
\SetWatermarkText{DRAFT}
\SetWatermarkScale{1.3}
\SetWatermarkLightness{0.9}

%-------------------------------------------------------------------------------------------
% end preamble setup, begin document
%-------------------------------------------------------------------------------------------
\begin{document}

	\title{\papertitle\footnote{Manuscript under review; see \url{http://geoffboeing.com/publications/multiscale-analysis-street-networks/}}}
	\date{\paperdate}
	\author[]{\myname \thanks{Email: \href{mailto:\myemail}{\myemail}}}
	\affil[]{\myaffiliation}

\maketitle

\begin{abstract}
OpenStreetMap offers an underexplored source of worldwide geospatial data useful to urban researchers. This study uses the OSMnx software to automatically download and analyze 27,000 US street networks from OpenStreetMap at metropolitan, municipal, and neighborhood scales---namely, every US city and town ($N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$). It presents wide-ranging empirical findings on US urban form and street network characteristics, emphasizing measures relevant to graph theory, urban design, and morphology such as structure, connectedness, density, centrality, and resilience. In the past, street network data acquisition and processing has been challenging and ad hoc. This study illustrates the use of OSMnx and OpenStreetMap to consistently conduct street network analysis with extremely large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. These street networks and measures data sets have been shared in a public repository for other researchers to use.
\end{abstract}

\vspace{0.25cm}

\begin{multicols}{2}

@@

On May 20, 1862, Abraham Lincoln signed the Homestead Act into law, making land across the United States' Midwest and Great Plains available for free to applicants \citep{porterfield_homestead_2005}. Under its auspices over the next 70 years, the federal government distributed 10\% of the entire US landmass to private owners in the form of 1.6 million homesteads \citep{lee_kansas_1979, sherraden_inclusion_2005}. New towns with gridiron street networks sprang up rapidly across the Great Plains and Midwest, due to both the prevailing urban design paradigm of the day and the standardized rectilinear town plats used repeatedly to lay out instant new cities \citep{southworth_streets_1997}. Through path dependence, the spatial signatures of these land use laws, design paradigms, and planning instruments can still be seen today in these cities' urban forms and street networks. Cross-sectional analysis of American urban form can reveal these artifacts and histories through street networks at metropolitan, municipal, and neighborhood scales.

Network analysis is a natural approach to the study of cities as complex systems \citep{masucci_random_2009}. The empirical literature on street networks is growing ever richer, but suffers from some limitations---discussed in detail in \citep{boeing_osmnx:_2017} and summarized here. First, sample sizes tend to be fairly small due to data availability, gathering, and processing constraints: most studies in this literature that conduct topological or metric analyses tend to have sample sizes ranging around 10 to 50 networks \citep{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, strano_urban_2013}, which may limit the generalizability and interpretability of findings. Second, reproducibility has been difficult when the dozens of decisions that go into analysis---such as spatial extents, topological simplification and correction, definitions of nodes and edges, etc.---are ad hoc or only partly reported \citep[e.g.,][]{porta_network_2006, strano_urban_2013}. Third, and related to the first two, studies frequently oversimplify to planar or undirected primal graphs for tractability \citep[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009}, or use dual graphs despite the loss of geographic, metric information \citep{batty_network_2005, jiang_integration_2002, ratti_space_2004, crucitti_centrality_2006, crucitti_centrality_2006-1}.

This study addresses these limitations by conducting a morphological analysis of urban street networks at multiple scales, with large sample sizes, with clearly defined network definitions and extents for reproducibility, and using nonplanar, directed graphs. In particular, it examines 27,000 urban street networks---represented as primal, nonplanar, weighted multidigraphs with possible self-loops---at multiple overlapping scales across the US, focusing on structure, connectedness, centrality, and resilience. It examines the street networks of every incorporated city and town (population size $N=19655$), census urbanized area ($N=497$), and Zillow-defined neighborhood ($N=6857$) in the US. To do so, it uses OSMnx\footnote{OSMnx is freely available online at \url{https://github.com/gboeing/osmnx}}---a new street network research toolkit \citep{boeing_osmnx:_2017}---to download, construct, and analyze these street networks at metropolitan, municipal, and neighborhood scales. These street networks and measures data sets have been compiled and shared in a public repository at the Harvard Dataverse\footnote{\label{note_data_repo}Data repository available online at \url{https://dataverse.harvard.edu/dataverse/osmnx-street-networks}} for other researchers to repurpose.

@@

A network (also called a \emph{graph} in mathematics) comprises a set of nodes connected to one another by a set of edges. Street networks can be conceptualized as primal, directed, nonplanar graphs. A \emph{primal} street network represents intersections as nodes and street segments as edges. A \emph{directed} network has directed edges: that is, edge $uv$ points one-way from node $u$ to node $v$, but there need not exist a reciprocal edge $vu$. A \emph{planar} network can be represented in two dimensions with its edges intersecting only at nodes \citep{viana_simplicity_2013, fischer_spatial_2014}. Most street networks are nonplanar---due to grade-separated expressways, overpasses, bridges, tunnels, etc.---but most quantitative studies of urban street networks represent them as planar \citep[e.g.,][]{buhl_topological_2006, cardillo_structural_2006, barthelemy_modeling_2008, masucci_random_2009, strano_urban_2013} for tractability because bridges and tunnels are uncommon in some cities. Planar graphs may reasonably model the street networks of old European town centers, but poorly model the street networks of modern autocentric cities like Los Angeles or Shanghai with many grade-separated expressways, bridges, and underpasses \citep{boeing_planarity_2018}.

@@

This study uses OSMnx to download, construct, correct, analyze, and visualize street networks at metropolitan, municipal, and neighborhood scales. OSMnx is a Python-based research tool that easily downloads OpenStreetMap data for any place name, address, or polygon in the world, then constructs it into a spatially-embedded graph-theoretic object for analysis and visualization \citep{boeing_osmnx:_2017,boeing_methods_2017}. OpenStreetMap is a collaborative worldwide mapping project that makes its spatial data available via various APIs \citep{corcoran_analysing_2013, jokar_arsanjani_openstreetmap_2015}. These data are of high quality and compare favorably to CIA World Factbook estimates and US Census TIGER/Line data \citep{haklay_how_2010, over_generating_2010, zielstra_comparative_2011, maron_how_2015, wu_improving_2005, frizzelle_importance_2009}. In 2007, OpenStreetMap imported the TIGER/Line roads (2005 vintage) and since then, many community-led corrections and improvements have been made \citep{willis_openstreetmap_2008}. Many of these additions go beyond TIGER/Line's scope, including passageways between buildings, footpaths through parks, bike routes, and detailed feature attributes such as finer-grained street classifiers, speed limits, etc.

To define the study sites and their spatial boundaries, we use three sets of geometries. The first defines the metropolitan-scale study sites using the 2016 TIGER/Line shapefile of US census bureau urban areas. Each census-defined urban area comprises a set of tracts that meet a minimum density threshold \citep{u.s._census_bureau_2010_2010}. We retain only the \emph{urbanized areas} subset of these data (i.e., areas with greater than 50,000 population), discarding the small \emph{urban clusters} subset. The second set of geometries defines our municipal-scale study sites using 51 separate TIGER/Line shapefiles (again, 2016) of US census bureau \emph{places} within all 50 states plus DC. We discard the subset of \emph{census-designated places} (i.e., small unincorporated communities) in these data, while retaining every US city and town. The third set of geometries defines the neighborhood-scale study sites using 42 separate shapefiles from Zillow, a real estate database company. These shapefiles contain neighborhood boundaries for major cities in 41 states plus DC. This fairly new data set comprises nearly 7,000 neighborhoods, but as \citet{schernthanner_spatial_2016} point out, Zillow does not publish the methodology used to construct these boundaries. However, despite its newness it already has a track record in the academic literature: \citet{besbris_effect_2015} use Zillow boundaries to examine neighborhood stigma and \citet{albrecht_indicator_2014} use them to study neighborhood-level poverty in New York.

For each of these geometries, we use OSMnx to download the (drivable, public) street network within it, a process described in detail in \citet{boeing_osmnx:_2017} and summarized here. First OSMnx buffers each geometry by 0.5 km, then downloads the OpenStreetMap \enquote{nodes} and \enquote{ways} within this buffer. Next it constructs a street network from these data, corrects the topology, calculates street counts per node, then truncates the network to the original, desired polygon. OSMnx saves each of these networks to disk as GraphML and shapefiles. Finally, it calculates metric and topological measures for each network, summarized below. Such measures extend the toolkit commonly used in urban form studies \citep{talen_measuring_2003, ewing_travel_2010}.

@@

Brief descriptions of these OSMnx-calculated measures are discussed here, but extended technical definitions and algorithms can be found in e.g. \citep{trudeau_introduction_1994, albert_statistical_2002, dorogovtsev_evolution_2002, brandes_network_2005, costa_characterization_2007, newman_structure_2003, newman_networks:_2010, barthelemy_spatial_2011, cranmer_navigating_2017}. The \emph{average street segment length} is a linear proxy for block size and specifies the network's grain. \emph{Node density} divides the node count by the network's area, while \emph{intersection density} excludes dead-ends to represent the density of street junctions. \emph{Edge density} divides the total directed network length by area, while \emph{street density} does the same for an undirected representation of the graph (to not double-count bidirectional streets). \emph{Average circuity} measures the ratio of edge lengths to the great-circle distances between the nodes these edges connect, indicating the street pattern's curvilinearity (cf. \citet{boeing_morphology_2018}).

The network's \emph{average node degree} quantifies connectedness in terms of the average number of edges incident to its nodes. The \emph{average streets per node} adapts this for physical form rather than directed circulation. It measures the average number of physical streets that emanate from each node (i.e., intersection or dead-end). The distribution and proportion of streets per node characterize the type, pervasiveness, and spatial dispersal of network connectedness and dead-ends. Connectivity represents the fewest number of nodes or edges that will disconnect the network if they are removed and is thus an indicator of resilience. A network's \emph{average node connectivity} (ANC)---the mean number of internally node-disjoint paths between each pair of nodes---more usefully represents how many nodes must be removed on average to disconnect a randomly selected pair of nodes \citep{beineke_average_2002,dankelmann_bounds_2003}. Brittle points of vulnerability characterize networks with low average connectivity.

A node's \emph{clustering coefficient} represents the ratio between its neighbors' links and the maximum number of links that could exist between them \citep{jiang_topological_2004, opsahl_clustering_2009}. The weighted clustering coefficient weights this by edge length and the average clustering coefficient is the mean of the clustering coefficients of all the nodes. \emph{Betweenness centrality} evaluates how many of the network's shortest paths pass through some node (or edge) to indicate its importance \citep{barthelemy_betweenness_2004,huang_trajgraph:_2016, zhong_revealing_2017}. A network's \emph{maximum betweenness centrality} (MBC) measures the share of shortest paths that pass through the network's most important node: higher maximum betweenness centralities suggest networks more prone to inefficiency if this important choke point should fail. Finally, \emph{PageRank} ranks nodes based on the structure of incoming links and the rank of the source node \citep{brin_anatomy_1998, jiang_predicting_2008, agryzkov_algorithm_2012, chin_geographically_2015, gleich_pagerank_2015}.

@@

\begin{table*}[htbp]

@@

\small

@@

\begin{figure*}[htbp]
	\centering
	\includegraphics[width=1\textwidth]{fig01.png}
	\caption{Intersection density and average streets per node per urbanized area in the contiguous US.}
	\label{fig:fig01}
\end{figure*}

\begin{figure*}[htbp]
	\centering
	\includegraphics[width=0.7\textwidth]{fig02.png}
	\caption{Distribution of node types in 9 urbanized areas, with number of streets emanating from the node on the $x$-axis and proportion of nodes of this type on the $y$-axis }
	\label{fig:fig02}
\end{figure*}

@@

@@

\end{multicols}

\begin{sidewaystable*}[p]

@@

\end{sidewaystable*}

\begin{multicols}{2}

@@

Table \ref{tab:measures_cities} presents summary statistics of street network characteristics across the entire data set of 19,655 cities and towns---every incorporated city and town in the US. Following recent work by \citet{barthelemy_modeling_2008} and \citet{strano_urban_2013}, we examine the relationship between the total street length $L$ and the number of nodes $n$ across different cities. The former proposed a model of city network evolution in which $L$ and $n$ scale nonlinearly as $n^{1/2}$, and the latter suggested that this relationship applies cross-sectionally, using a small empirical sample of ten European cities. However, the latter's small sample size may limit the generalizability of this finding. We examine the relationship between $L$ and $n$ across every US city and town and instead find a strong linear relationship ($r^{2}=0.98$), as depicted in Figure \ref{fig:fig03}. We also find a similar linear relationship at the metropolitan ($r^{2}=0.99$) and neighborhood ($r^{2}=0.98$) scales.

@@

Previous findings \citep[e.g.,][]{masucci_random_2009} suggest street segment lengths in an urban network follow a power-law distribution. We find that these networks instead generally follow lognormal-style right-skewed distributions. This makes theoretical sense as most street networks are not truly scale-free: for example, a typical street network might comprise very few very long street segments (e.g., 1 km), more medium-length segments (e.g., 250 m), many short segments (e.g., 80 m), but very few very short segments (e.g., 10 m). To test this in a statistically rigorous way, we fit a set of candidate distributions to the street segment lengths of each US city/town. The candidate distributions comprise the lognormal, Gumbel, gamma, exponentiated Weibull, Fréchet, power-law, uniform, and exponential distributions. We then assess each fit with a Kolmogorov-Smirnov test, which tells us how closely each fitted distribution matches the observed data. Power-law distributions provide the best fit for fewer than 0.7\% of these cities. In contrast, the exponentiated Weibull distribution provides the best fit 56\% of the time, followed by the Gumbel (19\%), gamma (11\%), and lognormal (7\%) distributions.

An exception to this general pattern, of course, lies in consistently-sized orthogonal grids filling a city's incorporated spatial extents. Such distributions are extremely peaked around a single value: the linear length of a grid block. We find that such cities are not uncommon in the US, particularly between the Mississippi River and the Rocky Mountains: the Great Plains states are characterized by a unique street network form that is both orthogonal and reasonably dense. The \emph{former} is partly the result of topography (flat terrain that allows idealized grids) and design history (rapid platting and development during the late nineteenth century) that favor orthogonal grids, as discussed earlier. The \emph{latter} results from the fact that most towns across the Great Plains exhibit minimal suburban sprawl. Thus, municipal boundaries snugly embrace the gridlike street network, without extending to accommodate a vast peripheral belt of twentieth century sprawl, circuity, and \enquote{loops and lollipops} \citep{southworth_streets_1997} that characterizes cities in e.g. California that were settled in the same era but later subjected to substantial suburbanization.

\begin{figure*}[htbp]
	\centering
	\includegraphics[width=1\textwidth]{fig03.png}
	\caption{The linear relationship between total street length and number of nodes in the street networks of every US urbanized area, city/town, and Zillow neighborhood.}
	\label{fig:fig03}
\end{figure*}

\begin{figure*}[htbp]
	\centering
	\includegraphics[width=1\textwidth]{fig04.png}
	\caption{Contiguous US states by median of mean streets per node and by median of mean street segment length in municipal street networks.}
	\label{fig:fig04}
\end{figure*}

\begin{table*}[htbp]

@@

\footnotesize

@@

@@

\begin{figure*}[htbp]
	\centering
	\includegraphics[width=0.8\textwidth]{fig05.png}
	\caption{Square-mile comparisons of central cities and their suburbs. Left: top, downtown Philadelphia, PA; bottom, its suburb, King of Prussia. Middle: top, downtown Portland, OR; bottom, its suburb, Beaverton. Right: top, downtown San Francisco, CA; bottom, its suburb, Concord.}
	\label{fig:fig05}
\end{figure*}

@@

To illustrate these morphologies, Figure \ref{fig:fig05} compares one square mile of the centers of Philadelphia, Portland, and San Francisco with one square mile of each of their suburbs. The connectedness and fine-grain of the central cities are clear, as are the disconnectivity and coarse-grain of their suburbs. In fact, the suburbs have more in common with one another---despite being hundreds or thousands of miles apart---than they do with their central city neighbors, suggesting that land use and an era's prevailing design paradigm is paramount to geographical localism and regional context. The top row of Figure \ref{fig:fig05} represents an era of planning and development that preceded the automobile, while the bottom row reflects the exclusionary zoning and mid--late twentieth century era of automobility in residential suburb design---namely the \enquote{loops and lollipops} and the \enquote{lollipops on a stick} design patterns \citep{southworth_streets_1997}.
